# Reproducible Research: Peer Assessment 1
The objective of this assignment is to evaluate data anonymously collected from 
a personal activity monitoring device. This data contains 17,568 samples 
measured at five minute intervals during the months of October and November 
2012. In addition, the [Peer Assessment assignment #1 description][1] includes 
the following definition of the variables contained in this data set:
* steps: Number of steps an individual takes during a five minute interval  
* date: "The date on which the measurement was taken in YYYY-MM-DD format"  
* interval: "Identifier for the 5-minute interval in which [the] measurement was taken"

The objective of the following code chunk is to setup the R language 
environment by performing four operations. First, this code chunk defining
a character vector (i.e. required.packages) that defines the required packages. 
The second operation performed by this R software is initializing charater 
vector (i.e. new.packages) based on a [comparison of required packages with the 
currently installed packages][2]. Third, this code chunk installs the required
packages by passing the new.packages character vector to the install.packages 
function. The fourth operation performed by this R code chunk is loading the 
following required packages:
* data.table  
* ggplot2  
* lubridate  
```{r setupEnvironment, message=FALSE}

# Verify required packages are installed
required.packages <- c("lubridate",
                       "data.table",
                       "ggplot2",
                       "xtable")

new.packages <- required.packages[!(required.packages %in% 
                                    installed.packages()[,"Package"])]

if(length(new.packages)) {
  install.packages(new.packages)
} 

# Load required packages
library(data.table)
library(ggplot2)
library(lubridate)
```

## Loading and preprocessing the data
The template [Github repository][1] for this assignment includes a compressed 
file (i.e activity.zip) that contains the personal activity monitoring data.
Therefore, the first step performed by the following code chunk is to determine
whether or not a "Data" directory has been created. If not, the R software 
contained in this if statement creates the "Data" directory and [extracts the
data from activity.zip][3]

```{r loadAndCleanData}
if (!file.exists("./Data")) {
    dir.create("./Data")    
    unzip("./activity.zip", exdir="Data")
}

activityData <- read.csv("./Data/activity.csv",
                         header=TRUE,
                         stringsAsFactors=FALSE)

activityData$date <- ymd(activityData$date)

originalIntervals <- unique(activityData$interval)

activityData$interval <- as.factor(activityData$interval)

levels(activityData$interval) <- seq(1,length(originalIntervals))

activityData <- data.table(activityData)

numberOfFiveMinuteIntervals <- 12 * 24

minOriginalIntervals <- min(originalIntervals)
maxOriginalIntervals <- max(originalIntervals)
```

There are `r numberOfFiveMinuteIntervals` 5 minute intervals in a day. However, 
the original intervals range from `r minOriginalIntervals` to 
`r maxOriginalIntervals`. Therefore, the purpose of the above code chunk  
is to:  
1. Convert the 5-minute interval variable into a factor variable in order to 
facilitate computing the average daily activity.  
2. Set the 5-minute intervals to correspond to the number of 5 minute intervals
in one twenty-four hour period.  

## What is mean total number of steps taken per day?
```{r totalNumberOfSteps, fig.height=4, fig.width=7, message=FALSE}
computeTotalNumberStepsPerDay <- function(activityData) {
    totalNumberStepsPerDay <- 
        as.data.frame(activityData[complete.cases(activityData),
                                   sum(steps),by=date])
    
    colnames(totalNumberStepsPerDay) <- c("date","totalnumberofsteps")
    
    return(totalNumberStepsPerDay)
}

totalNumberStepsPerDay <- computeTotalNumberStepsPerDay(activityData)

ggplot(totalNumberStepsPerDay, aes(x=totalnumberofsteps)) +
    geom_histogram(binwidth=1000,fill="white",colour="black") + 
    xlab("Total Number of Steps / Day")

ggsave("./figure/totalNumberOfSteps.png")

cat(sprintf("Mean total number of steps taken per day: %.1f", 
            mean(totalNumberStepsPerDay$totalnumberofsteps)),
    sprintf("Median total number of steps taken per day: %.1f", 
            median(totalNumberStepsPerDay$totalnumberofsteps)),
    fill=TRUE,
    sep="\n")
```

## What is the average daily activity pattern?
```{r dailyActivity, fig.height=4, message=FALSE}
dailyActivty <- as.data.frame(activityData[complete.cases(activityData),
                                           mean(steps),
                                           by=interval])

dailyActivty$interval <- as.numeric(dailyActivty$interval)
colnames(dailyActivty) <- c("interval","averagenumberofsteps")

ggplot(dailyActivty, aes(x=interval,y=averagenumberofsteps)) + 
    geom_line(size=1) +
    coord_cartesian(xlim=c(min(dailyActivty$interval),
                           max(dailyActivty$interval)),
                    ylim=c(min(dailyActivty$averagenumberofsteps),
                           max(dailyActivty$averagenumberofsteps))) +
    theme_gray(base_size=14) + 
    xlab("5 minute interval") + 
    ylab("Average Number of Steps") +
    ggtitle("Average Daily Activity Pattern") +
    scale_x_continuous(breaks=c(1,50,100,150,200,250))

maxInterval <- 
    as.integer(dailyActivty[which(dailyActivty$averagenumberofsteps == 
                                  max(dailyActivty$averagenumberofsteps)),
                            "interval"])

maxIntervalTime <- as.POSIXlt(paste(activityData[1,date],"00:00:00"))
maxIntervalTime$min <- maxIntervalTime$min + maxInterval*5
maxIntervalTime <- strftime(maxIntervalTime,"%H:%M (24 hour)")

ggsave("./figure/dailyActivity.png")
```

On average across all the days in the dataset, the five minute interval that 
contains the maximum number steps is #`r maxInterval` which corresponds to   
`r maxIntervalTime`.

## Imputing missing values
```{r computeNumberMissingValues}
totalNumberMissingValues <- nrow(activityData) - 
                            sum(complete.cases(activityData))
```

The total number of missing values is `r totalNumberMissingValues`, which is 
computed as by subtracting the number of rows where all of the variables (i.e. 
columns) are specified from the total number of rows.

```{r imputeMissingValues, fig.height=4, message=FALSE}
dailyActivty$interval <- as.factor(dailyActivty$interval)

for (n in seq(1,nrow(dailyActivty))) {
    intervalIndex <-
        which(activityData$interval == dailyActivty[n,"interval"])
    
    naIndex <- 
        intervalIndex[is.na(activityData[intervalIndex,steps])]
    
    avgIntegerNumberOfSteps <- 
        as.integer(round(dailyActivty[n,"averagenumberofsteps"]))
    
    activityData[naIndex,steps:=avgIntegerNumberOfSteps]
}

totalNumberStepsPerDay <- computeTotalNumberStepsPerDay(activityData)

ggplot(totalNumberStepsPerDay, aes(x=totalnumberofsteps)) +
    geom_histogram(binwidth=1000,fill="white",colour="black") + 
    xlab("Total Number of Steps / Day") + 
    ggtitle("NA's set to mean value")

ggsave("./figure/imputeMissingValues.png")

cat(sprintf("Mean total number of steps taken per day: %.1f", 
            mean(totalNumberStepsPerDay$totalnumberofsteps)),
    sprintf("Median total number of steps taken per day: %.1f", 
            median(totalNumberStepsPerDay$totalnumberofsteps)),
    fill=TRUE,
    sep="\n")
```

## Are there differences in activity patterns between weekdays and weekends?

```{r activityPatternDifferences, fig.height=4, message=FALSE, }
assign("depthtrigger", 60, data.table:::.global)
activityData[,daytype:=weekdays(date)]

weekday <- c("Monday",
             "Tuesday",
             "Wednesday",
             "Thursday",
             "Friday")

for (curDay in weekday) {
    activityData[which(activityData$daytype == curDay),daytype:="weekday"]
}

weekendDay <- c("Saturday",
                "Sunday")

for (curDay in weekendDay) {
    activityData[which(activityData$daytype == curDay),daytype:="weekend"]
}

activityData$daytype <- as.factor(activityData$daytype)

activityData$interval <- as.numeric(activityData$interval)

activityPattern <- as.data.frame(activityData[,mean(steps),by=list(daytype,interval)])

colnames(activityPattern) <- c("daytype","interval","avgnumberofsteps")

ggplot(activityPattern,aes(x=interval,y=avgnumberofsteps)) + 
    geom_line(size=1) + 
    theme_gray(base_size=14) +
    facet_grid(daytype ~ .) + 
    facet_wrap(~daytype,nrow=2) + 
    xlab("Interval") + ylab("Number of Steps")
```

[1]: https://github.com/rdpeng/RepData_PeerAssessment1
[2]: http://stackoverflow.com/questions/9341635/how-can-i-check-for-installed-
r-packages-before-running-install-packages
[3]: http://www.r-bloggers.com/read-compressed-zip-files-in-r/